# Similarities and differences among models of productivity {#prod-comp}

## Introduction

One of the central challenges facing a child learning language is when and how
to form generalizations beyond their exact input. An English-learning child
needs to infer, for example, that not only do many verbs that they have heard
form the past tense by adding +/\textipa{d}/ to the stem (*talk* $\rightarrow$
*talked*, *bark* $\rightarrow$ *barked*) but that novel forms can do so as well
(*glip* $\rightarrow$ *glipped*). At the same time, the child needs to encode
that there are exceptions to this pattern and learn their form (*go*
$\rightarrow$ *went*). The question of how learners come to know which units in
their language generalize under what conditions in the face of ambiguous,
exception-riddled evidence can be called the *problem of productivity*
[@odonnell2015].

While the problem of productivity has been at the heart of the study of language
for centuries [@bauer2001], there is an increasing consensus that the
distribution of linguistic units -- that is, its pattern of usage of words,
stems, affixes, etc. across different contexts -- plays a central role
[@baayen2009; @bauer2005; @odonnell2015; @yang2016]. In particular, three recent
theories -- Baayen's hapax-based approach [@baayen1993; @baayen2001;
@baayen2009], Yang's tolerance principle [@yang2005; @yang2010; @yang2016], and
O'Donnell's productivity and reuse tradeoff [@odonnell2011; @odonnell2015;
@odonnell2009] -- are distinguished by their theoretical foundations,
mathematical precision, empirical coverage, and contact with linguistic theory.

The theories make fundamentally different psychological assumptions in terms of
whether cognitive computations are performed sequentially or in parallel and
whether linguistic representations are probabilistic or deterministic. These
differences represent opposing views about the nature of linguistic computation.
Thus, a crucial scientific goal is to isolate and study linguistic phenomena in
which these different approaches lead to differing empirical predictions. Our
work aims to examine these three leading theories of morphological productivity
and delineate both the substantive similarities and the crucial differences
among their assumptions and predictions.

To perform this evaluation, we study variants of these models for which
inferences are directly comparable in an artificial domain that captures the
essential features of the problem of productivity. We perform a series of
analyses that show that the models make systematically differing predictions.
Specifically, we examine three dimensions for which we predict there to be
informative differences between these models: number of types, token counts and
irregular ranks, and rule cost.

In the rest of this paper, we first lay out the problem of productivity as
formalized in this work. We then delineate the representational and
distributional assumptions and properties of each model. Next, we describe the
the three dimensions listed above in terms of the background for their
importance and the predictions we have about their effects on the models. We
then specify the designs of our simulations and show their results. Finally, we
conclude by connecting our model comparisons to future empirical investigations
that will enable adjudication between the theories.


## Setup

### Learning problem

Consider a given morphological system, for example, the much-studied English
past tense. Verbs are inflected to form the past tense using a variety of
morphological processes. For the purposes of this work, we refer to any such
process as a *transformation*, a function that takes a stem and returns its
inflected form. With respect to a certain transformation (e.g. \textsc{stem}
$\rightarrow$ \textsc{stem} + /\\textipa{d}/), some verbs are *regular* (their
inflected form matches the application of the transformation, e.g. *use*
$\rightarrow$ *used*) and some are *irregular* (their inflected form does not
match the application of the transformation, e.g. *go* $\rightarrow$ *went,* not
*go* $\rightarrow$ *goed*).

To isolate the problem of productivity, we presuppose that the learner has
matched up bases with their inflected forms and isolated the potentially
productive transformation, i.e. we set aside the problem of rule discovery and
focus on productivity inference. The task for any learner or model is then to
determine for each transformation, whether to use it productively, i.e. whether
to apply it to novel stems. For example, consider a toy corpus of the English
past tense:

```{=tex}
\vspace{1em}
\begin{tabular}{llll}
\hline
\textbf{verb} & \textbf{stem} & \textbf{transformation} & \textbf{count} \\
\hline
% go & \textsc{stem} $\rightarrow$ went & $c_\text{go}$ \\
% walk & \textsc{stem} $\rightarrow$ \textsc{stem} + /d/ & $c_\text{walk}$ \\
% run & \textsc{stem} $\rightarrow$ ran & $c_\text{run}$ \\
% use & \textsc{stem} $\rightarrow$ \textsc{stem} + /d/ & $c_\text{use}$ \\
% bark & \textsc{stem} $\rightarrow$ \textsc{stem} + /d/ & $c_\text{bark}$ \\
go & /\textipa{gow}/ & \textsc{stem} $\rightarrow$ \textipa{wEnt} & $c_\text{go}$ \\
walk & /\textipa{wOk}/ & \textsc{stem} $\rightarrow$ \textsc{stem} + /\textipa{d}/ & $c_\text{walk}$ \\
run & /\textipa{\*r2n}/ & \textsc{stem} $\rightarrow$ \textipa{\*raen} & $c_\text{run}$ \\
use & /\textipa{juz}/ & \textsc{stem} $\rightarrow$ \textsc{stem} + /\textipa{d}/ & $c_\text{use}$ \\
play & /\textipa{plej}/ & \textsc{stem} $\rightarrow$ \textsc{stem} + /\textipa{d}/ & $c_\text{play}$ \\
\hline
\end{tabular}
\vspace{1em}
```

Faced with this input, the learner would need to determine whether to treat the
transformation \textsc{stem} $\rightarrow$ \textsc{stem} + /\\textipa{d}/ as
productive and memorize the forms *go/went* and *run/ran*, or to not treat it as
productive and memorize all five forms.

Inferences about productivity could in principle take into account many
distributional properties of the linguistic input. Does it matter how many
regular and irregular verbs there are? Does it matter how many times each verb
occurs in the corpus? Does it matter if irregulars tend to be more frequent than
regulars, or *vice versa*? Theories of productivity have proposed various
distributional information sources as the determinants of productivity,
including type frequencies [@bybee1995], token frequencies, single occurrence
frequencies [@baayen1993; @baayen2001; @baayen2009], number of exceptions
[@yang2005; @yang2010; @yang2016], and functions of one or more of the above
[@aronoff1976; @barnwell2010; @taatgen2002; @odonnell2015].

### Measuring productivity

We define the productivity of a transformation under some model as the
probability of using that transformation productively, i.e. applying it to a
novel stem. In this view, a model can be thought of as a function that takes a
corpus as input and returns a probability distribution over transformations
given a novel stem. For each model, we define its productivity measure
$\mathcal{P}^*$ as the probability it assigns to using a target transformation
$\tau$ given a novel stem $s^*$ and a corpus $D$:
$$\mathcal{P}^* = P(\tau\, |\, s^*,D)$$

We will be defining this quantity for each model under consideration and
computing it under various regimes in our simulations. Any time we refer to the
"productivity" of a model, we are referring to the corresponding
$\mathcal{P}^*$.

## Models

In this section, we give an overview of the assumptions of each model and
illustrate its inferences using the toy corpus shown in
Figure \ref{fig:examples}. We focus on translating the underlying machinery of
each framework to a common language, aiming to make clearer the similarities and
differences among the models and pinpoint the ways in which their predictions
differ. For each model, we define its productivity measure $\mathcal{P}^*$.

Throughout, we use the notation of the input to a model being a corpus $D$ of
triplets stem, transformation, count: $(s_i,t_i,c_i)$, where there are $n$ stems
(in the set of stems $S = E \cup R$) of which $k$ are irregular (in $E$) and
$n-e$ are regular (in $R$). Additionally, we designate the counts of the
irregulars as $c^E_1, c^E_2, \ldots, c^E_e$ and the counts of the regulars as
$c^R_1, c^R_2, \ldots, c^R_{n-e}$.

\input{04-prod-comp/representations}

### Hapax-based statistical estimators

By far the most thoroughly studied and widely used measures of productivity are
Baayen's P\*\*, P\* and P [@baayen2001]. Because productive affixes frequently
give rise to new words, their word frequency distributions typically contain a
large number of low frequency forms. This observation suggests that a reasonable
estimate of the productivity of an affix might be based on the number or
proportion of low-frequency words associated with the affix. Baayen's statistics
are based on this idea. In particular, all three quantities estimate the
probability of unobserved (i.e., novel) words using the proportion of word types
which appear only once in a corpus (hapaxes). This intuition can be rigorously
justified under several different sets of mathematical assumptions
[@baayen2001]. Baayen's P\* (*expanding productivity*) corresponds to a direct
estimator of our $\mathcal{P}^*$.

In particular, for a corpus of $N$ words, let $|H$\| be the number of words in
the corpus with frequency 1 (hapaxes): $H = \{s \in S\ |\ c_s = 1\}$; and let
$|H_R|$ be the number of hapaxes in the corpus that use the target
transformation: $H_R = \{s \in R\ |\ c_s = 1\}$. P\* is then the proportion of
all hapaxes in the corpus that use the target transformation:

$$ \mathcal{P}^*_\texttt{HAPAX} = \frac{|H_R|}{|H|}$$

### Tolerance principle

In a series of studies, Yang has developed a theory of productivity known as the
tolerance principle [@yang2005; @yang2010; @yang2016]. Below we describe first
the machinery underlying the tolerance principle, the elsewhere condition serial
search model, and then its analytic form. Our model comparison set includes both
versions.

#### Elsewhere condition serial search

The elsewhere condition serial search (`ECSS`) framework posits that inflecting
a stem is based on serial search through a frequency-ranked list of forms. If
the stem isn't a found in the list, an "elsewhere condition" indicates how to
inflect it. For a given transformation, productivity inferences arise from
comparing two candidate lists, one in which all forms are listed (i.e. stored)
vs. one in which only the irregulars are listed and the regulars fall under the
elsewhere condition. These two hypotheses are compared based on their expected
processing times. For the example corpus, Figure \ref{fig:example_nonprod} shows
the full listing (no rule) representation, where all stems are stored in
descending order of frequency, so the most frequent stem (*go*) requires one
time step, the second most frequent stem (*walk*) requires two time steps, and
so on. Figure \ref{fig:example_prod} shows the rule representation, where the
regulars are pulled out of this list, so they all require a fixed number of
steps (in this case, 3).

More concretely, for our set $S$ of $n$ stem types, let
$T = t_1, t_2, \ldots, t_n$ be the processing time of each stem. The probability
(normalized token frequency) of a stem $s$ is given by $P(s)$, its frequency
rank among all stems is given by $r_s$, and for exceptions, their frequency rank
among exceptions is given by $r_e$.

The no rule hypothesis corresponds to the search time of each stem being equal
to its rank among all stems. So the expected processing time $T_\text{NR}$
averaged over all stems is given by:

$$ \mathbb{E}[T_\text{NR}] = \sum_{s \in S} P(s)t_s = \sum_{s \in S} P(s)r_s $$

For the rule hypothesis, the processing times of exceptions are equal to their
ranks within the list of exceptions $t_s = r_e$, while each of the regulars
requires the same number of processing time steps, $k$, plus the fixed cost
$\omega$ of applying the rule $t_s = e + \omega$. So the expected processing
time $T_\text{R}$ is given by:

$$ \mathbb{E}[T_\text{R}] = \sum_{s \in S} P(s)t_s = \sum_{s \in S} P(s) (\delta_e r_e + (1 - \delta_e)(k + \omega)) $$

Where $\delta_e$ is an indicator variable that is equal to one if $s$ is an
exception.

Inference of productivity then comes from comparing the two expected running
times $\mathbb{E}[T_\text{R}]$ and $\mathbb{E}[T_\text{NR}]$.

There are multiple possibilities for how this hypothesis selection might be used
to determine $\mathcal{P}^*$. Given a novel stem, the model could categorically
apply the regular transformation if and only if the rule hypothesis is selected:

$$
\begin{aligned}
\mathcal{P}^*_{\texttt{ECSS}}
&=
  \begin{cases}
    1 & \mathbb{E}[T_\text{R}] < \mathbb{E}[T_\text{NR}] \\
    0 & \text{otherwise}
  \end{cases}
\end{aligned}
$$

Or it could always apply the regular transformation under the rule hypothesis
and select a transformation uniformly at random under the no rule hypothesis:

$$
\begin{aligned}
\mathcal{P}^*_{\texttt{ECSS}}
&=
  \begin{cases}
    1 & \mathbb{E}[T_\text{R}] < \mathbb{E}[T_\text{NR}] \\
    \frac{1}{k+1} & \text{otherwise}
  \end{cases}\\[2ex]
\end{aligned}
$$

Here we use the latter $\mathcal{P}^*$ for `ECSS`, in line with [@schuler2016].

#### Analytic tolerance principle {#tp}

Yang makes a number of distributional assumptions that allow for the derivation
of a generally-stated evaluation metric that does not depend on token
frequencies. This simplification requires assuming that:

(1) Token frequencies follow the Zipfian distribution (for a stem of rank $r_s$,
    $P(s)=\frac{1}{r_sH_n}$ where $H_n = \sum_{i=1}^n \frac{1}{i}$.

(2) Irregulars are a random subset of all stems (each stem's probability of
    being irregular is $k/n$).

(3) A random sample from a Zipfian distribution is also Zipfian, so the token
    frequencies of the irregulars also follow the Zipfian distribution. (for a
    stem of rank $r_e$ among exceptions, $P(s)=\frac{1}{r_eH_e}$.

(4) There is no cost associated with rule application ($\omega=0$).

With these assumptions about the input, the model yields the threshold that if
$k$ types are irregular out of $n$ total, the transformation will be considered
productive if and only if $k \leq \frac{n}{\log{n}}$. As such, the threshold
depends only on the type frequencies of regulars and irregulars and does not
interact in any way with their token frequencies.

Under the analytic tolerance principle threshold `TP`, if there is a productive
transformation, it will always be used for novel stems. It is less clear how
this model ought to treat novel stems if there is not a productive
transformation. As above for `ECSS`, we use a fallback to picking a
transformation uniformly at random:

$$
\mathcal{P}^*_{\texttt{TP}} =
  \begin{cases}
    1 & e \leq \frac{k}{\log{k}} \\
    \frac{1}{e+1} & \text{otherwise}
  \end{cases}
$$

### Productivity and reuse tradeoff

Another theory of productivity is O'Donnell's productivity and reuse tradeoff
[@odonnell2011; @odonnell2015; @odonnell2009]. This approach is based on the
idea that the productivity of linguistic units can be determined by optimizing a
probabilistic trade-off between a pressure to store fewer, more reusable
primitive units and a pressure to account for each linguistic expression in as
few computational steps as possible. These pressures are quantified in terms of
probability and competition is resolved in parallel. To capture this tradeoff,
the productivity and reuse framework makes the following two assumptions.

First, it assumes that the probability of an expression is the product of the
probabilities associated with the use of each lexical item in the derivation of
that expression expression. Since probabilities are numbers between $0$ and $1$,
a product of probabilities declines rapidly (geometrically quickly) as the
number of probabilities in the product increases. The more complex a derivation
is, therefore, the smaller its probability will be, on average. Of course, the
exact probability will depend on the probabilities of individual lexical items
in a derivation. If all of the lexical items are very probable, then the
expression will also be more probable. Nevertheless, more complex derivations
will tend to have lower probabilities regardless of the specific lexical items
within them. This assumption creates a bias towards derivations which use fewer
lexical items per expression, in the extreme case perhaps treating entire
expressions as single lexical units.

Second, it assumes that the probability of a lexical item in some context in a
derivation is proportional to the number of times that lexical item was used in
the same context in the derivation of other expressions. This implements a
rich-get-richer scheme: The probability that a lexical item be used will
increase as it is used more. For a given corpus, the fewer items in the lexicon,
the more often (on average) each item will have to be used across derivations
and, thus, the more probable each item will be. Relating the probability of
lexical items to their frequency of use straightforwardly implements the bias
towards smaller lexicons with smaller, more reusable lexical items.

These assumptions lead to a tradeoff. One one hand, the learner could avoid
complex derivation by storing a large number of relatively big lexical items.
While this would minimize the complexity of derivations, each lexical item would
be less reusable across expressions, and thus less probable. On the other hand,
the learner could favor a small number of relatively small lexical items. These
lexical items would be highly reusable across contexts and thus highly probable.
However, many of them would be required to analyze each expression. The best
lexicon is the one which optimizes these competing constraints.

As a consequence of this evaluation metric, productivity is determined by a
complex interplay between type and token frequencies. More regular types would
generally lead to the regular transformation being more likely to be productive
(more forms can be decomposed, meaning more savings from having a table for the
regular transformation). However, this is not necessarily the case, and
interacts with the relative frequencies of the regular and irregular forms. High
token frequency regular forms can actually inhibit the drive toward
productivity, since a high probability of use creates a higher computational
cost. The biggest driver of productivity is a high number of low frequency
regular forms, corresponding to the intuition that the best evidence for the
productivity of a transformation is seeing it used productively, i.e. applied to
novel forms.

In the general case, a full model based on the productivity and reuse tradeoff
would consider every combination of storage and computation for each form. This
results in a huge hypothesis space and complex inference problem that can be
solved by a stochastic search procedure. Here we take a different approach by
applying the simplifying assumptions described above to the underlying
representations used by the framework. For a given corpus, all forms are
initially treated as computed and are iteratively stored until further storage
does not increase the model's posterior score. This corresponds to a search
through the posterior space that allows for possibilities where some (high
frequency) regulars are stored and/or irregulars computed. The model's
parameters are set to correspond to uninformative priors over storage states.

The productivity metric for this model is then the probability of using the
target transformation given a novel stem, conditioned on the storage state found
by the procedure described above.

## Predictions

We lay out three key dimension for which we predict the models that we are
comparing to have informative convergences and divergences in their behaviors.
In our simulations, we will quantitatively evaluate these predictions.

### Number of types

__Background__. How do productivity inferences about a specific transformation
depend on the overall number of types to which that transformation can apply?
Are transformations with smaller or larger domains able to tolerate more
exceptions while being productive? For the tolerance principle, the productivity
threshold, being $\frac{n}{\log(n)}$, grows sublinearly as function of $n$, and
so "'[S]maller' rules... can tolerate a relatively higher number of exceptions.
Large rules are more vulnerable" [@yang2016, p. 66]. According to Yang, this
property is "the most surprising, and in my view the most important,
quantitative aspect of the Tolerance Principle" [@yang2016, p. 66].

For both productivity and reuse and hapax-conditioned productivity, there is no
straightforwardly necessary relationship between productivity and number of
types, because their productivity inferences depend not just on the overall
number of types but on further properties of the type and token distribution of
regulars and irregulars. But there is a likely relationship such that as you
increase the number of types, under reasonable assumptions about word
frequencies, you are more likely to get more regulars that drive towards
productivity (i.e. more regular hapaxes for HC, more non-stored regulars for
PR).

__Predictions__. As such, we expect tolerance to number of irregulars as a
function of number of types to _decrease_ for `TP` and to _increase_ for `PR`
and `HC`, though possibly in different ways depending on properties of the
corpus setup. For `ECSS`, to the extent that the more general elsewhere
condition serial search model is well approximated by the assumptions of the
analytic tolerance principle, it should exhibit the same behavior.

### Token counts and irregular ranks

__Background__. A key feature of the tolerance principle is its simplicity --
only the numbers of regular and irregular types need to be taken into account,
rather than more complex corpus statistics. However, this property comes about
from making a number of assumptions that allow the full `ECSS` model to be
simplified into the `TP` analytic form (as described in Section\ \ref{tp}). For
the `ECSS` measure of productivity, token counts and irregular ranks matter --
they translate into probabilities that figure into the calculation of expected
processing time. But for `TP`, it is assumed that irregulars are a random subset
of all types and that the token count distributions for all types and for
irregulars specifically are both Zipfian. Yang posits that the `TP` analytic
form has the necessary explanatory power without necessitating the `ECSS`
calculation and without dependence on token counts. He also argues that a
"[e]ven if all the exceptions are in the top half of the $N$ items, the
threshold... will hardly move" [@yang2016, p. 65].

For both productivity and reuse and hapax-conditioned productivity, token counts
and irregular ranks are crucial. For `HC` they determine how many regulars are
in the tail of the word frequency distribution such that they have a count of 1;
for `PR` they similarly influence how many regulars have low counts and are thus
provide pressure against storage of the regular transformation.

The dependence on token counts and irregular ranks is particularly important to
evaluate given that the relation between frequency and irregularity is an
empirical question. Yang notes that while "it has often been observed that
exceptions tend to clustered the [sic] high-frequency region of words... the
concentration of English irregular verbs at the very top appears to be an
outlier in a wide range of empirical cases" [@yang2016, p. 65]. In recent work,
@wu2019 confirmed statistically in a sample of 28 languages the long-standing
observation that higher frequency forms tend to be irregular and that irregular
forms tend to be higher frequency.

__Predictions__. For `TP`, only type counts matter and token counts should be
irrelevant. For `ECSS`, in cases where the derivational assumptions of `TP` are
satisfied (i.e. irregulars are random), its behavior should be similar to that
of `TP`, but when they are altered its behavior may diverge. In contrast, for
`PR` and `HC`, token counts and irregulars ranks should both play a role, such
that corpus setups where there are more low frequency regulars should be more
conducive to productivity.

### Rule cost

__Background__. The focus of the theory of productivity and reuse is on an
optimization tradeoff between storage and computation, in that there is both a
cost associated with storing items (fewer and more reusable stored units is
better) and a cost associated with applying transformations (fewer computation
steps is better). `ECSS` also implements an optimization (over expected
processing time), but there is no cost associated with applying a
transformation. However, the underlying machinery behind `ECSS` includes not
just this "rule selection" process but also a "rule application" process. Rule
application "is sensitive to the frequency of the morphological constituent"
[@yang2016, p. 55], much like `PR` is sensitive to item frequency. The question
becomes, how does the behavior of `ECSS` change when there is a non-zero cost
associated with the application of a transformation? How does that relate to the
behavior of `PR`?

Hapax-conditioned productivity is a statistical estimator, rather than a
cognitive model.  As such, it doesn’t include a notion of storage or computation
and is omitted from this analysis.

__Predictions__. We predict that the addition of rule cost will change behavior
of `ECSS`, such that it behaves more similarly to `PR` when the rule cost is
higher.


## Analyses

The goal of our analyses is to investigate how inferences about productivity are
determined by various distributional properties of the input, and to
quantitatively evaluate the predictions described above. To perform this
evaluation, we (1) construct a space of parametrically-varying input corpora,
(2) define two simulation regimes using those corpora, and (3) conduct analyses
using these simulations to investigate the role of each of the three factors
described above.

### Corpora

We consider corpora with one potentially productive target transformation that
applies to a subset of stems. The input to a model is a corpus that consists of
a set of triplets <`stem`, `transformation`, `count`>, which gives the count of
the number of times the form given by the combination of that stem and
transformation appears in the corpus.

For all corpora we use a Zipfian distribution of token frequencies, such that
for a stem $s$ with frequency rank $r_s$:

$$ P(s) = \frac{1}{r_s H_n}, H_n = \sum_{i = 1}^n \frac{1}{i} $$

These artificially generated corpora are determined by:

-   Number of types ($n$).

-   Number of irregular types (each irregular form is assigned to a unique
    transformation, and all other forms are assigned to the target
    transformation, so $k$ irregulars corresponds to $k + 1$ transformations).

-   Placement of irregulars among all types in frequency rank. Irregular ranks
    are either a random sample of all ranks or are determined such that the gaps
    between them are given by the Zipfian distribution. As an example, a corpus
    with $n=10$ and $k=4$ would have ranks (from first to last) given by
    `IrIrrIrrrIrrrr` (where `I` indicates an irregular and `r` indicates a
    regular).

```{r}
source("04-prod-comp/analysis_params.R")
```

### Simulations

Using this space of corpus parametrizations, we run a set of simulations that
examine the models' behavior under various regimes. We use two types of
simulations, one for computing for productivity values and one for finding
productivity thresholds.

#### Productivity values

We generate corpora where $n$ is fixed to `r fixed_n`, and $k$ ranges from 0 to
`r max_prop`$\cdot n$. For each irregular ranking method, we directly compute
the $\mathcal{P}^*$ of each model and examine its value as a function of number
of $k$.

#### Productivity thresholds

We define the target transformation to be productive under a given model if its
$\mathcal{P}^*>0.5$. For a given $n$, we define a model's productivity threshold
as the greatest number of irregular types such that the target transformation is
productive. We examine how these productivity thresholds change a function of
$n$ under each irregular ranking method.

The total number of types ranges between `r threshold_min_n` and
`r threshold_max_n`. For each value of $n$, the possible number of irregulars
ranges from 0 to `r max_prop`$\cdot n$. For a given value of $n$, we find the
productivity threshold for each model by finding the number of irregulars $k$
such that the model gives productive for $k$ and non-productive for $k+1$. For
`TP` and `ECSS` this consists of checking every number of irregulars from 0 up
until the model gives non-productive; for `PR` this consists of doing binary
search over the range of possible numbers of irregulars.

## Results

### Analysis 1: Number of types

Across all different simulations, there is a striking difference in the shapes
of productivity threshold curves between `PR` and `TP`, as shown in
Figure \ref{fig:thresholds-ranks}. As the overall number of types increases,
`PR` can tolerate an *increasing* proportion of irregulars, while `TP` can
tolerate a *decreasing* proportion of irregulars. `HC` tolerates a roughly
constant proportion of irregulars. An immediate consequence of this observation
is that under `TP`, it is challenging to account for morphological paradigms
with a generalization that applies to a minority of types (or even to a majority
less than the `TP` threshold), but is nevertheless productive. This is famously
the case in the German plural, where there is a transformation that applies to a
tiny minority of nouns, but is used productively. One way to attempt to resolve
this issue is to break down types into subsets that only certain transformations
can apply to, leaving a smaller set over which to determine the productivity of
the minority generalization [@yang2016].

```{r thresholds-ranks, fig.cap="For each model, curves show the maximum proportion of irregulars for which the regular transformation is inferred to be productive, as a function of total number of types. Panels show different irregular ranking methods."}
knitr::include_graphics("04-prod-comp/figures/thresholds-ranks-1.pdf")
# thresholds_plot |>
#   filter(group != "ECSS" | param == 0) |>
#   plot_thresholds()
```

```{r fixed, fig.cap="For each model, curves show the productivity of the regular transformation as a function of number of irregulars out of a fixed number of types. Panels show different irregular ranking methods."}
knitr::include_graphics("04-prod-comp/figures/fixed-1.pdf")
# fixed_data_plot |>
#   filter(num_exceptions <= 0.6 * num_stems) |>
#   plot_fixed_data()
```

### Analysis 2: Token frequencies and irregular ranks

__Productivity thresholds__. Figure \ref{fig:thresholds-ranks} also shows the
productivity threshold curves under various exception ranking regimes, with
irregulars ranked uniformly in the left panel and irregulars more frequent in
the right panel.

The tolerance curve for `PR` is sensitive to the positions of irregulars, in
that `PR` can tolerate fewer irregulars when they are more frequent than when
they are uniform. However this difference is small, and may be an artifact of
our particular simulation setup.

This analysis reveals a difference in behavior between `TP` and `ECSS`. When
irregulars are ranked uniformly, `ECSS` approximates `TP` closely. As irregulars
are skewed more towards the high end of the frequency distribution, however,
`ECSS` systematically deviates from `TP`, and if irregulars are sufficiently
highly frequent, the productive representation of the data is strictly faster in
processing time that the full listing representation, so the transformation is
*always* treated as productive.

__Productivity values__. Figure \ref{fig:fixed} shows the productivity
values under various ranking regimes, with irregulars ranked uniformly in the
left panel and irregulars more frequent in the right panel.

Under both irregular ranking methods, the curves for `PR` and `HC` are broadly
similar, with a gradual decrease in productivity as the number of irregulars
increases. There is a larger gap between the two when irregulars are uniform as
opposed to when they are more frequent. This is consistent with the idea that
for both `PR` and `HC`, productivity is driven primarily by a large number of
low-frequency regulars (directly as hapaxes for `HC`, indirectly as pressure for
a reusable unit for `PR`). Their productivity trajectories are thus broadly
similar, but differ less when the irregulars are more frequent, in that the
increasing irregulars don't affect the tail of regulars as much.

(ref:demotraj-cap) For `TP` and each rule application cost of `ECSS`, curves
show the maximum proportion of irregulars for which the regular transformation
is inferred to be productive, as a function of total number of types.
```{r thresholds-ecss-tp, out.width="50%", fig.cap="(ref:demotraj-cap)"}
knitr::include_graphics("04-prod-comp/figures/thresholds-ecss-tp-1.pdf")
# thresholds_plot |>
#   filter(group %in% c("TP", "ECSS"), str_detect(dist_name, "uniform")) |>
#   plot_thresholds(0.4)
```

### Analysis 3: Rule cost

__Productivity thresholds__. In Figure \ref{fig:thresholds-ecss-tp}, we show the
productivity threshold curves for `TP` and for `ECSS` with rule costs between 0
and `r max(threshold_rule_costs)`. As the rule cost increases, the behavior of
`ECSS` becomes less similar to `TP` and more similar to `PR`, with an increasing
proportion of exceptions tolerated as a function of overall number of types.
`ECSS` without a rule cost, along with `TP`, encodes a pressure *for*
productivity, in that combining regulars into an elsewhere condition can reduce
their expected search time. The addition of a rule cost adds a pressure
*against* productivity, in that the addition of the elsewhere condition incurs a
separate processing cost. As such the `ECSS` model can be thought of as encoding
a trade-off between storage and computation similar to that of `PR`. However,
the threshold curves for `ECSS` appear to asymptote at a much lower proportion
of irregulars than for `PR`.

__Productivity values__. Referring back to the productivity values in Figure
\ref{fig:fixed}, we again see a convergence between the behavior of `TP` and
`ECSS` when irregulars are ranked uniformly and for a rule cost of 0, and a
divergence in other settings. Further, we see why in the threshold analysis,
`ECSS` is always productive when irregulars are frequent, as that is equivalent
to $\mathcal{P}^*_{\texttt{ECSS-sig}}>0.5$. However, even when the irregulars
are more frequent, the productivity trajectories of `ECSS` can be differentiated
by rule cost -- the costlier it is to apply the rule, the less productive `ECSS`
tends to be. `ECSS` with a non-zero rule cost again more closely resembles `PR`
than `TP`, presumably because, as discussed above, it encodes a version of
storage-computation tradeoff.

## Conclusion {#prod-comp-conclusion}

We compared and contrasted the assumptions and behaviors of three leading models
of morphological productivity -- the tolerance principle, the theory of
productivity and reuse, and hapax-based estimation. To do this, we generated a
set of parametrically-varying corpora and examined how properties of this
inputs affected the patterns in productivity inferences made by each model.

First, we found a fundamental difference between productivity and reuse on one
hand and the tolerance principle and the more general elsewhere condition
serial search on the other hand. As the overall number of types that a
morphological transformation could apply to increases, `PR` can tolerate an
_increasing_ proportion of irregular types, while `TP` and `ECSS` can tolerate
a _decreasing_ proportion of irregular types. Commonly studied morphological
systems, for example the English past tense, therefore can't distinguish between
the predictions of these models, because they have a large number of types and
a very low proportion of irregulars.

Second, we isolated several ways in which the analytical form of the tolerance
principle systematically differs from the `ECSS` model from which it's derived.
The derivation of the tolerance principle crucially involves the assumption
that irregular types a random sample of all types, and for inputs where this
assumption doesn't hold (i.e. irregulars are skewed towards being higher
frequency), `ECSS` becomes increasingly prone to productivity.

Additionally, elsewhere condition serial search lends itself to the addition of
a rule cost component, reflecting a tradeoff between storage and computation
rather than an optimization based solely on storage cost. An increasing rule
cost drastically changes the behavior of `ECSS`, becoming more similar to `PR`.

An important note about this work is in all of these analyses, the exact values
returned by the models should not be treated as quantitative predictions -- the
input corpora are simulated rather than empirical and the parameter values are
set uninformatively rather than tuned to specific data. Rather, the shapes of
models' predictions within a simulation paradigm gives signatures of their
behavior that can be compared with one another.

Future work could involve finding morphological systems or experimental
paradigms that would enable evaluation of the model differences we isolated
here.
